Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-01-12 EvolutionaryRescue1.tex
copy-editing

There has been extensive theoretical analysis of evolutionary rescue.
\citet{orr2008population,orr2014population} thoroughly investigated the simplest case in which there is a specific single mutation or low-frequency variant that is sufficient to rescue the population. 
\citet{martin2013probability} extended these results to consider the effects of a range of rescue genotypes that can vary in both growth rate and variance in offspring number.
\citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
\citet{osmond_genetic_2020} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model.

@@

Here, we study evolutionary rescue after a sudden environmental change caused by initiating anti-cancer drug treatment. 
Essentially, this consists of combining known results from the evolutionary rescue literature, focusing on the quantities and parameter values that are relevant for cancer.
Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
While our focus is on the application of evolutionary rescue theory to evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available.
We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}; note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}. %change 47
We estimate the effect of aneuploidy on the tumor's evolutionary rescue probability. When aneuploidy provides drug resistance, it can directly rescue the tumor. However, when aneuploidy provides tolerance to the drug, it may act as an evolutionary ``stepping stone'' or ``springboard'' \citep{Yona2015,martin2013probability,osmond_genetic_2020}, delaying extinction, and thereby allowing the tumor more time to acquire a resistance mutation on top of the aneuploid background. 
As mentioned above, evidence suggests that aneuploidy may be a common strategy for tumor adaptation to drug therapy. Still, it is unknown how often aneuploidy provides tolerance and acts as a stepping stone.
We also estimate the mean time until a tumor cell population reaches its pre-treatment size following drug therapy. 
Given that aneuploidy is present in many tumors before the onset of therapy \citep{lukow2021chromosomal,ben2020context}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics. Additionally, we are interested in the timescale of evolutionary rescue and the impact that aneuploidy has on the time necessary for the tumor to overcome drug therapy.

@@

\paragraph{Parameterization.}
To parametrize the simulations, we assume that the cells under consideration are melanoma cells and rely on \citet{rew2000cell} and \citet{bozic2013evolutionary} for the division and death rates, respectively. 
\citet{rew2000cell} report \emph{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double, disregarding cell death) for a large set of cancer types. The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day. We select this to be the division rate for sensitive and mutant cells. This is a reasonable assumption given that we assume that the mutation makes the cancer cells immune to the chemotherapeutic drugs. %change 40

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Standing vs. \emph{de-novo} genetic variation.}

@@

Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by de-novo variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_s}$ is replaced by the cost of aneuploidy $c$, such that
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{c}{\abs{\Delta_s}}.
\end{equation}
This result has been previously reported by \citet{orr2008population} for the one-step evolutionary rescue and by \citet{martin2013probability} for two-step evolutionary rescue. % change 42
Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations are accurate (\Cref{rescue_denovo}). 
Standing genetic variation will drive evolutionary rescue if sensitive cells die rapidly (growth rate $\Delta_s$ is very negative) due to a strong effect of the drug on sensitive cells or if the cost of aneuploidy in the drug-free environment, $c$, is small.  
In contrast, \emph{de-novo} aneuploid cells will have a greater contribution to rescue if the cost of aneuploidy, $c$, is large, the effect of the drug on sensitive cells is weak ($\Delta_s$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_s=0.1$, $\mu_s=0.14$, $u=10^{-2}$, $\tilde{u}=10^{-3}$, and $c=0.07$, the ratio of the threshold tumor sizes for standing vs. \emph{de-novo} variation is $\tilde{N}^*_a/N^*_a \approx 17.5$, which means that \emph{de-novo} genetic variation is  the main driver of evolutionary rescue.

@@

We hypothesized that standing genetic variation (the existence of aneuploid cancer cells in the tumor before the onset of therapy) could facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. We found that a drug that reduces the sensitive growth rate and does not significantly increase the chromosome missegregation rate will likely lead to evolutionary rescue through standing genetic variation (\Cref{rescue_denovo,eq:Nsgv}).
Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f \gg u \lambda_s / \abs{\Delta_s} \approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather than through \emph{de-novo} aneuploidy. However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore, we expect evolutionary rescue to occur by \emph{de-novo} aneuploidy.

@@

\begin{figure}
\begin{subfigure}{0.5\textwidth}
A\\
\includegraphics[width=1\textwidth]{Figures/RatiodsPlot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
B\\
\includegraphics[width=1\textwidth]{Figures/ratio_uPlot.pdf}
\end{subfigure}
\caption{
\textbf{Standing genetic variation facilitates the evolutionary rescue of cancer.}
(A)  Ratio of threshold tumor sizes for rescue by standing genetic variation and by de-novo variation, $\tilde{N}_a^*/N_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, as a function of the sensitive growth rate $\Delta_s$.  Standing genetic variation will drive adaptation to the drug if the sensitive population is rapidly declining ($\Delta_s\ll0$) due to a stronger effect of the drug on sensitive cells. Red dots represent simulation results, and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G). Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3},u=10^{-2}, v=10^{-7}$.
(B) Ratio of threshold tumor size $\tilde{N}_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, and $N_a^*$ as a function of the ratio of aneuploidy rates $\tilde{u}/u$. \emph{De-novo} aneuploids will have a larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$). Red dots represent simulation results, and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G). Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3}, v=10^{-7}$.
}
\label{rescue_denovo}
\end{figure}
%%%%%%%%%%%

>>>>

There has been extensive theoretical analysis of evolutionary rescue.
\citet{orr2008population,orr2014population} thoroughly investigated the simplest case in which there is a specific single mutation or low-frequency variant that is sufficient to rescue the population. 
\citet{martin2013probability} extended these results to consider the effects of a range of rescue genotypes that can vary in both growth rate and variance in offspring number.
\citet{uecker2016role} modeled a situation in which \textit{two} mutations at different loci (with recombination between them) are necessary for rescue. 
\citet{osmond_genetic_2020} considered a population that must achieve rescue, potentially via multiple mutations, by adapting on a fitness landscape described by Fisher's geometric model,
and found that for some parameter values rescue by two mutational steps is more likely than rescue by a single step.

@@

Here, we study evolutionary rescue after a sudden environmental change caused by initiating anti-cancer drug treatment. 
Essentially, this consists of combining known results from the evolutionary rescue literature, focusing on the quantities and parameter values that are relevant for cancer.
Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
While our focus is on the application of evolutionary rescue theory to evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available, similar to part of \citet{osmond_genetic_2020} but focused on a different fitness landscape.

We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}; note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}. %change 47
We estimate the effect of aneuploidy on the tumor's evolutionary rescue probability. When aneuploidy provides drug resistance, it can directly rescue the tumor. However, when aneuploidy provides tolerance to the drug, it may act as an evolutionary ``stepping stone'' or ``springboard'' \citep{Yona2015,martin2013probability,osmond_genetic_2020}, delaying extinction, and thereby allowing the tumor more time to acquire a resistance mutation on top of the aneuploid background. 
As mentioned above, evidence suggests that aneuploidy may be a common strategy for tumor adaptation to drug therapy. Still, it is unknown how often aneuploidy provides tolerance and acts as a stepping stone.
We also estimate the mean time until a tumor cell population reaches its pre-treatment size following drug therapy. 
Given that aneuploidy is present in many tumors before the onset of therapy \citep{lukow2021chromosomal,ben2020context}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics. Additionally, we are interested in the timescale of evolutionary rescue and the impact that aneuploidy has on the time necessary for the tumor to overcome drug therapy.

@@

\paragraph{Parameterization.}
To parametrize the simulations, we assume that the cells under consideration are melanoma cells and rely on \citet{rew2000cell} and \citet{bozic2013evolutionary} for the division and death rates, respectively. 
\citet{rew2000cell} report \textit{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double, disregarding cell death) for a large set of cancer types. The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day. We select this to be the division rate for sensitive and mutant cells. This is a reasonable assumption given that we assume that the mutation makes the cancer cells immune to the chemotherapeutic drugs. %change 40

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Standing vs. \textit{de novo} genetic variation.}

@@

Therefore, the threshold tumor size for rescue by standing generation variation, $\tilde{N}^*_{a}$, is similar to the threshold for rescue by \textit{de novo} variation, $N^*_a$, except that the sensitive growth rate $\abs{\Delta_s}$ is replaced by the cost of aneuploidy $c$, such that
\begin{equation}\label{eq:Nsgv}
\frac{\tilde{N}^*_{a}}{N^*_{a}} = \frac{u}{\tilde{u}} \; \frac{c}{\abs{\Delta_s}}.
\end{equation}
This result has been previously reported by \citet{orr2008population} for the one-step evolutionary rescue and by \citet{martin2013probability} for two-step evolutionary rescue. % change 42
Comparing this approximation of $\tilde{N}^*_{a}/N^*_{a}$ to results of stochastic simulations, we find that the approximations are accurate (\Cref{rescue_denovo}). 
Standing genetic variation will drive evolutionary rescue if sensitive cells die rapidly (growth rate $\Delta_s$ is very negative) due to a strong effect of the drug on sensitive cells or if the cost of aneuploidy in the drug-free environment, $c$, is small.  
In contrast, \textit{de novo} aneuploid cells will have a greater contribution to rescue if the cost of aneuploidy, $c$, is large, the effect of the drug on sensitive cells is weak ($\Delta_s$ is close to zero), or if the drug induces the appearance of aneuploid cells ($u > \tilde u$).
For example, with  $\lambda_s=0.1$, $\mu_s=0.14$, $u=10^{-2}$, $\tilde{u}=10^{-3}$, and $c=0.07$, the ratio of the threshold tumor sizes for standing vs. \textit{de novo} variation is $\tilde{N}^*_a/N^*_a \approx 17.5$, which means that \textit{de novo} genetic variation is  the main driver of evolutionary rescue.

@@

We hypothesized that standing genetic variation (the existence of aneuploid cancer cells in the tumor before the onset of therapy) could facilitate evolutionary rescue by reducing the waiting time for the appearance of aneuploid cells. We found that a drug that reduces the sensitive growth rate and does not significantly increase the chromosome missegregation rate will likely lead to evolutionary rescue through standing genetic variation (\Cref{rescue_denovo,eq:Nsgv}).
Furthermore, if the fraction of tumor cells that have the beneficial aneuploidy is $f \gg u \lambda_s / \abs{\Delta_s} \approx2.5\%$, then evolutionary rescue is more likely to occur via standing variation rather than through \textit{de novo} aneuploidy. 
However, for the parameter values we focus on in our examples (\Cref{table1}), this fraction is an order of magnitude lower, and therefore, we expect evolutionary rescue to occur by \textit{de novo} aneuploidy.

@@

\begin{figure}
\begin{subfigure}{0.5\textwidth}
A\\
\includegraphics[width=1\textwidth]{Figures/RatiodsPlot.pdf}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
B\\
\includegraphics[width=1\textwidth]{Figures/ratio_uPlot.pdf}
\end{subfigure}
\caption{
\textbf{Standing genetic variation facilitates the evolutionary rescue of cancer.}
(A)  Ratio of threshold tumor sizes for rescue by standing genetic variation and by \textit{de novo} variation, $\tilde{N}_a^*/N_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, as a function of the sensitive growth rate $\Delta_s$.  Standing genetic variation will drive adaptation to the drug if the sensitive population is rapidly declining ($\Delta_s\ll0$) due to a stronger effect of the drug on sensitive cells. Red dots represent simulation results, and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G). Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3},u=10^{-2}, v=10^{-7}$.
(B) Ratio of threshold tumor size $\tilde{N}_a^*$, when a fraction $\frac{\tilde{u}\lambda_s}{c}$ is aneuploid at the start of treatment, and $N_a^*$ as a function of the ratio of aneuploidy rates $\tilde{u}/u$. \textit{De novo} aneuploids will have a larger contribution to the appearance of drug resistance if the drug induces the appearance of aneuploid cells ($u \gg \tilde u$). Red dots represent simulation results, and the error bars represent the $95\%$ confidence intervals obtained with bootstrap (Appendix G). Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09,\tilde{u}=10^{-3}, v=10^{-7}$.
}
\label{rescue_denovo}
\end{figure}
%%%%%%%%%%%
